---
title: "Final Project"
author: "David Robinson, Aditi Jackson, and Jonathan Joyner"
date: "December 6th, 2023"
output:
  pdf_document:
    latex_engine: pdflatex
  html_document:
    df_print: paged
always_allow_html: true
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory - EDE-Jackson_Joyner_Robinson
setwd("C:/ENVIRON 872/EDE_Jackson_Joyner_Robinson")
getwd()

# Load your packages
#install.packages(readxl)
#install.packages(tidyverse)
#install.packages(dplyr)
#install.packages(lubridate)
#install.packages(ggplot2)
#install.packages(sf)
#install.packages(mapview)
#install.packages(trend)
#install.packages(zoo)
#install.packages(kendall)
#install.packages(tseries)
#install.packages("corrplot")
#install.packages(leaflet)

library(readxl)
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(sf)
library(mapview)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
library(corrplot)
library(leaflet)

# Set your ggplot theme

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

# Load your datasets
eGRID2010 <- read_excel("Data/Raw/eGRID2010_Data.xls", 
    sheet = "PLNT10", skip = 4)
eGRID2012 <- read_excel("Data/Raw/eGRID2012_Data.xlsx", 
    sheet = "PLNT12", skip = 4)
eGRID2014 <- read_excel("Data/Raw/eGRID2014_Data.xlsx", 
    sheet = "PLNT14", skip = 1)
eGRID2016 <- read_excel("Data/Raw/egrid2016_Data.xlsx", 
                      sheet = "PLNT16", skip = 1)
eGRID2018 <- read_excel("Data/Raw/eGRID2018_Data.xlsx",
                      sheet = "PLNT18", skip = 1)
eGRID2020 <- read_excel("Data/Raw/eGRID2020_Data.xlsx",
                      sheet = "PLNT20", skip = 1)
Gov_Data <- read.csv("Data/Raw/US_Governors_1775_2020.csv")

States_shapefile_shp <- st_read("Data/Raw/us-state-boundaries.shp")


```
# Rationale and Research Questions

To conduct a preliminary exploration of the ways in which the political systems and the electricity grid coexist, our team has formulated the following research questions:

  1) What is the relationship between the percent penetration of renewable technology (i.e.,     solar, wind) and state governor (using the political affiliation of the state governor as     a proxy for a state’s political leanings) in a given year / over a period of time?
  
  2) Has renewable penetration grown in any states that have had a single-party governor over     a longer period of time?

# Dataset Information
This dataset was prepared in Fall 2023 as a final project for Environmental Data Analytics (ENV 872) at Duke University's Nicholas School of the Environment.

The repository contains 10 years of data (2010-2020) detailing the United State's renewable energy generation capacity by governing state political party. There are three main categories for our dataset. They are:
1) eGRID plant data with coordinates
2) US state governor data by party and year
3) State geographic boundaries shapefile

Table: Data Structure Summary - eGRID

Detail  | Description
:------------- | -------------:
Data Source  | EPA
Retrieved from  | https://www.epa.gov/egrid/download-data
Variables Used  | PNAME, PSTATABB, NAMEPCAP, PLFUELCT, LAT, LON
Data Range | 2010 - 2020

Table: Data Structure Summary - Governors

Detail  | Description
:------------- | -------------:
Data Source  | University of Pennsylvania
Retrieved from  | https://www.openicpsr.org/openicpsr/project/102000/version/V3/view?path=/openicpsr/102000/fcr:versions/V3/united_states_governors_1775_2020.csv&type=file
Variables Used  | governor, state, time_in_office, party, year
Data Range | 1775 - 2020

\newpage

# Wrangling eGRID

```{r Wrangle eGRID}
# selecting data for desired columns from imported eGRID datasets

#Note column "PLPFGNCT" for 2010 -- note that "PLFUELCT" was dropped and replaced
eGRID2010_sub <- eGRID2010 %>%
  mutate(YEAR="2010") %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLPFGNCT,NAMEPCAP,LAT,LON)

eGRID2010_sub$PLFUELCT <- eGRID2010_sub$PLPFGNCT
eGRID2010_sub$PLPFGNCT <- NULL

#Note column "PLPFGNCT" for 2012 -- note that "PLFUELCT" was dropped and replaced
eGRID2012_sub <- eGRID2012 %>%
  mutate(YEAR="2012") %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLPFGNCT,NAMEPCAP,LAT,LON)

eGRID2012_sub$PLFUELCT <- eGRID2012_sub$PLPFGNCT
eGRID2012_sub$PLPFGNCT <- NULL

eGRID2014_sub <- eGRID2014 %>%
  mutate(YEAR="2014") %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLFUELCT,NAMEPCAP,LAT,LON)

eGRID2016_sub <- eGRID2016 %>%
  mutate(YEAR="2016") %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLFUELCT,NAMEPCAP,LAT,LON)

eGRID2018_sub <- eGRID2018 %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLFUELCT,NAMEPCAP,LAT,LON)

eGRID2020_sub <- eGRID2020 %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLFUELCT,NAMEPCAP,LAT,LON)

# merging eGRID data, filtering for renewable fuels
eGRID_2010_2020 <- rbind(eGRID2010_sub,eGRID2012_sub,eGRID2014_sub,eGRID2016_sub,
      eGRID2018_sub,eGRID2020_sub)

# checking fuel types
unique(eGRID_2010_2020$PLFUELCT)

# filtering for desired fuel types
eGRID_2010_2020_RENEW <- filter(eGRID_2010_2020,PLFUELCT=="SOLAR"|PLFUELCT=="WIND"|PLFUELCT=="GEOTHERMAL"|
          PLFUELCT=="HYDRO"|PLFUELCT=="BIOMASS"|PLFUELCT=="NUCLEAR")

#Change year column to character
eGRID_2010_2020_RENEW$YEAR <- as.character(eGRID_2010_2020_RENEW$YEAR)

# checking fuel types were filtered correctly
unique(eGRID_2010_2020_RENEW$PLFUELCT)

# checking to ensure all years of interest are present
unique(eGRID_2010_2020_RENEW$YEAR)

# saving process data as CSV
write_excel_csv(eGRID_2010_2020_RENEW, 
                path = "Data/Processed/eGRID_2010_2020_RENEW.csv")
```
\newpage

#Wrangling Gov Data

```{r Wrangling Gov Data, include=FALSE}

#Filtering governor data for 2010 to 2020
Gov_Data_10_20 <- filter(Gov_Data,year >= 2010)

#Creating State abbreviation column for gov data
Gov_Data_State <- Gov_Data_10_20 %>%
  mutate(StateAbbreviation = state.abb[match(state, state.name)])

#Changing year to character
Gov_Data_State <- Gov_Data_State %>%
  mutate(year = as.character(year))

#Separating time in office column
#Key assumption is that end year is the year to be used for analysis of political
#Rationale -- incumbent governor is in office for the majority of this year
Gov_Data_State <- Gov_Data_State %>%
  separate(time_in_office, into = c("Start_Year", "End_Year"), sep = " - ")

# filtering for columns of interest
States_shapefile_shp_filtered <- States_shapefile_shp %>%
  select(basename, stusab, geometry)

#Merging governor data with state shape file
GOV_States <- left_join(Gov_Data_State, States_shapefile_shp_filtered, by = c("StateAbbreviation" = "stusab"))

#Joining gov data with eGRID processed data
GOV_eGRID_10_20 <- left_join(eGRID_2010_2020_RENEW, Gov_Data_State, by = c("PSTATABB" = "StateAbbreviation", "YEAR" = "year"))

# dropping N/A's
GOV_eGRID_10_20 <- GOV_eGRID_10_20 %>%
  drop_na(LAT)

#Checking years of final dataset
unique(GOV_eGRID_10_20$YEAR)

#Checking fuel types of final dataset
unique(GOV_eGRID_10_20$PLFUELCT)

#Dataset information -- REVISIT THIS
dim(GOV_eGRID_10_20)
colnames(GOV_eGRID_10_20)

#Adding Final Processed file to processed folder
write_csv(GOV_eGRID_10_20, 
                path = "Data/Processed/GOV_eGRID_10_20.csv")

```

\newpage

# Exploratory Analysis 
```{r Exploratory Time Series Analysis, include=FALSE, fig.retina=2, dev='pdf'}

#Exploratory distribution analysis using box, point, and violin plot

ggplot(GOV_eGRID_10_20, aes(x = YEAR, y = NAMEPCAP, color= PLFUELCT)) +
  geom_point() +
  ylim(0, 5000)

ggplot(GOV_eGRID_10_20, aes(x = YEAR, y = NAMEPCAP)) +
  geom_boxplot() +
  ylim(0, 50)

ggplot(GOV_eGRID_10_20, aes(x = YEAR, y = NAMEPCAP)) +
  geom_violin() +
  ylim(0, 50)

Republican_Renewables <- GOV_eGRID_10_20 %>%
  filter(party == "Republican") %>%
  ggplot(aes(x = YEAR, y = NAMEPCAP, color = PLFUELCT)) +
  geom_point() +
  ylim(0, 5000)
plot(Republican_Renewables)

Democrat_Renewables <- GOV_eGRID_10_20 %>%
  filter(party == "Democrat") %>%
  ggplot(aes(x = YEAR, y = NAMEPCAP, color = PLFUELCT)) +
  geom_point() +
  ylim(0, 5000)
plot(Democrat_Renewables)

```

\newpage

# Analysis

```{r Regression Analysis, include=FALSE}

#WRANGLING
# creating subset of governor data to sum no. of democrats in office by state
Years_in_Office <- Gov_Data_State %>%
  mutate(dem_count = ifelse(party=="Democrat",1,0),
         rep_count=ifelse(party=="Republican",1,0))%>%
  group_by(StateAbbreviation)%>%
  summarise(total_dem_count = sum(dem_count),total_rep_count=sum(rep_count))

# aggregating nameplate capacity by state
GOV_eGRID_10_20_TOTAL_CAPACITY <- GOV_eGRID_10_20 %>%
  group_by(PSTATABB) %>%
  summarize(TOTAL_NAMEPCAP = sum(NAMEPCAP))

  
# joining datasets to create data set for regression analysis
Capacity_by_Party <-left_join(Years_in_Office,GOV_eGRID_10_20_TOTAL_CAPACITY,by = c("StateAbbreviation"="PSTATABB"))
                               
#SINGLE REGRESSION
# running simple linear regression of Democratic party on nameplate capacity
Regression_DemParty_by_Capacity <- lm(Capacity_by_Party$TOTAL_NAMEPCAP~Capacity_by_Party$total_dem_count)
summary(Regression_DemParty_by_Capacity)

Regression_RepParty_by_Capacity <- lm(Capacity_by_Party$TOTAL_NAMEPCAP~Capacity_by_Party$total_rep_count)
summary(Regression_RepParty_by_Capacity)

# plotting single regressions - capacity by Dem
ggplot(Capacity_by_Party, aes(x=total_dem_count, y=TOTAL_NAMEPCAP)) +
  geom_point()

# plotting single regressions - capacity by Rep
ggplot(Capacity_by_Party, aes(x=total_rep_count, y=TOTAL_NAMEPCAP)) +
  geom_point()

# results
#The x variable is "number of years a Democratic governor was in office" and the y variable is "Total nameplate capacity from renewable energy sources".
# Ho: There is no relationship between renewable capacity and the number of years a Democratic governor was in office.
#Given the R-squared value, around virtually none of the variability in total nameplate capacity from renewable energy sources is explained by the number of years a Democratic governor was in office. Given the high p-value of 0.9239, we fail to reject H0. Thus, there is no relationship between renewable capacity and the number of years a Democratic governor was in office. (Linear regression; df = 46, R-squared = 0.0002003, F-statistic = 0.009217, p-value = 0.9239)
#Note that we can this single regression can also be applied to "# of years a Republican governor was in office" -- assuming the party is binary between Democrat and Republican, this should drive a similar regression result. Indeed, that is the case as evidenced by the analogous output for Republican governors.
#See the below plots of these regressions which help to numerical outputs and write-ups above.

#MULTIPLE REGRESSION
# trying multiple linear regression
MP_Reg_Capacity_by_Party <- lm(data=Capacity_by_Party,TOTAL_NAMEPCAP~total_dem_count+total_rep_count)

summary(MP_Reg_Capacity_by_Party)

# results
#The predictor variables are "number of years a Democratic governor was in office" and "# of years a Republican governor was in office" and the y variable is "Total nameplate capacity from renewable energy sources".
#Given the R-squared value, virtually none of the variability in total nameplate capacity from renewable energy sources is explained by the number of years a Democratic governor was in office and the number of years that a Republican governor was in office. Given the high p-value of 0.9776, we fail to reject H0. Thus, there is no relationship between renewable capacity and the number of years a Democratic governor was in office and the number of years a Republican governor was in office. 
#(Multiple regression; df = 45, R-squared = 0.001006, F-statistic = 0.02266, p-value = 0.9776)

#PEARSON CORRELATION
# running a correlation test
cor.test(Capacity_by_Party$TOTAL_NAMEPCAP,Capacity_by_Party$total_dem_count)

# results
#The x variable is "number of years a Democratic governor was in office" and the y variable is "Total nameplate capacity from renewable energy sources".
# The calculated correlation coefficient between the two variables is approximately -0.014, which suggests a very weak relationship between the variables x and y. Moreover, the high p-value means that we fail to reject the null hypothesis (that there is no correlation). Thus, it appears that there is no significant correlation between the variables. (Pearson correlation; df = 46, Correlation coefficient = -0.01415372, p-value = 0.9239, 95% confidence interval ranges from -0.2970936 to 0.2710711))

#ANOVA
#ANOVA Test for party correlative effect on nameplate capacity
ANOVA_Capacity <- aov(data = Capacity_by_Party, TOTAL_NAMEPCAP ~ total_dem_count)
summary(ANOVA_Capacity)

# results
# Similar to the results above, the high p-value of 0.924 indicates that the relationship between "number of years a Democratic governor was in office" and "total nameplate capacity between from renewable energy sources" is not statistically significant(ANOVA; df = 1, F-value = 0.009, p-value = 0.924,).

#TIME SERIES PLOT

GOV_eGRID_10_20_aggregate <- GOV_eGRID_10_20 %>%
  group_by(YEAR) %>%
  summarize(NAMEPCAPSUM = sum(NAMEPCAP, na.rm = TRUE))

GOV_eGRID.yearly.ts <- ts(GOV_eGRID_10_20_aggregate$NAMEPCAPSUM,
                          start=c(2010,1),
                          end = c(2020,1),
                          deltat = 2)
plot(GOV_eGRID.yearly.ts, 
     main = "Yearly Nameplate Capacity Sum", 
     xlab = "Year", 
     ylab = "Nameplate Capacity Sum")

#MANN-KENDALL TEST
GOV_eGRID.yearly.ts.mk <- Kendall::MannKendall(GOV_eGRID.yearly.ts)
summary(GOV_eGRID.yearly.ts.mk)

# results
# Given a p-value of less than 0.05 at 0.0085349, we reject H0 and conclude that there is a monotonic, yearly trend to the data.  Because tau is positive, we conclude that the trend is increasing (Mann-Kendall; tau = 1, p-value = 0.0085349)


```
\newpage

# Geospatial Analysis

```{r Geospatial Analysis, include=FALSE}

#EXPLORATORY
#Viewing State Boundaries
mapview(States_shapefile_shp)

#NC GEOSPATIAL ANALYSIS
#Viewing NC power plants in three different years
Gov_eGRID_NC_2012.sf <- GOV_eGRID_10_20 %>% 
  st_as_sf(coords = c('LON','LAT'), crs=4326) %>%
  filter(PSTATABB == "NC") %>%
  filter(YEAR == 2010:2012)

mapview(Gov_eGRID_NC_2012.sf, zcol="PLFUELCT")

Gov_eGRID_NC_2016.sf <- GOV_eGRID_10_20 %>% 
  st_as_sf(coords = c('LON','LAT'), crs=4326) %>%
  filter(PSTATABB == "NC") %>%
  filter(YEAR == 2010:2016)

mapview(Gov_eGRID_NC_2016.sf, zcol="PLFUELCT")

Gov_eGRID_NC_2020.sf <- GOV_eGRID_10_20 %>% 
  st_as_sf(coords = c('LON','LAT'), crs=4326) %>%
  filter(PSTATABB == "NC") %>%
  filter(YEAR == 2010:2020)

mapview(Gov_eGRID_NC_2020.sf, zcol="PLFUELCT")

#Combine spatial layers for NC
#Filter NC State Boundary by party
GOV_States.sf <- st_as_sf(GOV_States) %>%
  filter(StateAbbreviation == "NC")

#Layering NC Boundaries and plants for 2012

party_map <- mapview(GOV_States.sf, col.regions = "blue") +
  mapview(Gov_eGRID_NC_2012.sf, zcol="PLFUELCT")
print(party_map)

#Layering NC Boundaries and plants for 2016

party_map <- mapview(GOV_States.sf, col.regions = "red") +
  mapview(Gov_eGRID_NC_2016.sf, zcol="PLFUELCT")
print(party_map)

#Layering NC Boundaries and plants for 2016
party_map <- mapview(GOV_States.sf, col.regions = "blue") +
  mapview(Gov_eGRID_NC_2020.sf, zcol="PLFUELCT")
print(party_map)

#TX GEOSPATIAL ANALYSIS
#Viewing TX power plants in three different years
Gov_eGRID_TX_2012.sf <- GOV_eGRID_10_20 %>% 
  st_as_sf(coords = c('LON','LAT'), crs=4326) %>%
  filter(PSTATABB == "TX") %>%
  filter(YEAR == 2012)

mapview(Gov_eGRID_TX_2012.sf, zcol="PLFUELCT")

Gov_eGRID_TX_2020.sf <- GOV_eGRID_10_20 %>% 
  st_as_sf(coords = c('LON','LAT'), crs=4326) %>%
  filter(PSTATABB == "TX") %>%
  filter(YEAR == 2012:2020)

mapview(Gov_eGRID_TX_2020.sf, zcol="PLFUELCT")

#Combine spatial layers for TX
#Filter NC State Boundary by party
GOV_States.sf <- st_as_sf(GOV_States) %>%
  filter(StateAbbreviation == "TX")

#Layering TX Boundaries and plants for 2012

party_map <- mapview(GOV_States.sf, col.regions = "red") +
  mapview(Gov_eGRID_TX_2012.sf, zcol="PLFUELCT")
print(party_map)

#Layering TX Boundaries and plants for 2016
party_map <- mapview(GOV_States.sf, col.regions = "red") +
  mapview(Gov_eGRID_TX_2020.sf, zcol="PLFUELCT")
print(party_map)

#CA GEOSPATIAL ANALYSIS
#Viewing CA power plants in three different years
Gov_eGRID_CA_2012.sf <- GOV_eGRID_10_20 %>% 
  st_as_sf(coords = c('LON','LAT'), crs=4326) %>%
  filter(PSTATABB == "CA") %>%
  filter(YEAR == 2010:2012)

mapview(Gov_eGRID_CA_2012.sf, zcol="PLFUELCT")

Gov_eGRID_CA_2020.sf <- GOV_eGRID_10_20 %>% 
  st_as_sf(coords = c('LON','LAT'), crs=4326) %>%
  filter(PSTATABB == "CA") %>%
  filter(YEAR == 2012:2020)

mapview(Gov_eGRID_CA_2020.sf, zcol="PLFUELCT")

#Combine spatial layers for TX
#Filter CA State Boundary by party
GOV_States.sf <- st_as_sf(GOV_States) %>%
  filter(StateAbbreviation == "CA")

#Layering CA Boundaries and plants for 2012

party_map <- mapview(GOV_States.sf, col.regions = "blue") +
  mapview(Gov_eGRID_CA_2012.sf, zcol="PLFUELCT")
print(party_map)

#Layering CA Boundaries and plants for 2016
party_map <- mapview(GOV_States.sf, col.regions = "blue") +
  mapview(Gov_eGRID_CA_2020.sf, zcol="PLFUELCT")
print(party_map)
```

\newpage

## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 



\newpage

# Summary and Conclusions
#In summary, as it relates to the first research question, there is no apparent relationship #between the percent penetration of renewable energy technologies and state governor over the #period from 2010 - 2020. 
#There are a number of important caveats to make here -- renewable energy development #depends on many factors ranging from federal incentives to state-level renewable portfolio #standards to local zoning ordinances and regulations. Moreover, the dynamics of power markets #/ grid operators (e.g., investor-owned / vertically integrated / non-competitive markets such as #in the southeast versus competitive markets such as ERCOT in Texas) are critically influential, #particularly for the process of interconnection.
#While this project has led to some preliminary, high-level conclusions, it leads to a number of questions for further exploration: 
#	1. What is the magnitude of changes in renewables in the three example states of California (Democratic governor for the entire time period of interest), Texas (Republican governor for the entire time period of interest), and North Carolina (both Democratic and Republican governors for the time period of interest)?
#	2. How do local (e.g., country-level) politics influence renewable energy generation by sub-region of states?


\newpage

# References
<add references here if relevant, otherwise delete this section> 
