---
title: "Final Project"
author: "David Robinson, Aditi Jackson, and Jonathan Joyner"
date: "December 6th, 2023"
output:
  pdf_document:
    latex_engine: pdflatex
  html_document:
    df_print: paged
always_allow_html: true
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory - EDE-Jackson_Joyner_Robinson
setwd("C:/ENVIRON 872/EDE_Jackson_Joyner_Robinson")
getwd()

# Load your packages
#install.packages(readxl)
#install.packages(tidyverse)
#install.packages(dplyr)
#install.packages(lubridate)
#install.packages(ggplot2)
#install.packages(sf)
#install.packages(mapview)
#install.packages(trend)
#install.packages(zoo)
#install.packages(kendall)
#install.packages(tseries)
#install.packages("corrplot")
#install.packages(leaflet)

library(readxl)
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(sf)
library(mapview)
library(trend)
library(zoo)
library(Kendall)
library(tseries)
library(corrplot)
library(leaflet)

# Set your ggplot theme

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

# Load your datasets
eGRID2010 <- read_excel("Data/Raw/eGRID2010_Data.xls", 
    sheet = "PLNT10", skip = 4)
eGRID2012 <- read_excel("Data/Raw/eGRID2012_Data.xlsx", 
    sheet = "PLNT12", skip = 4)
eGRID2014 <- read_excel("Data/Raw/eGRID2014_Data.xlsx", 
    sheet = "PLNT14", skip = 1)
eGRID2016 <- read_excel("Data/Raw/egrid2016_Data.xlsx", 
                      sheet = "PLNT16", skip = 1)
eGRID2018 <- read_excel("Data/Raw/eGRID2018_Data.xlsx",
                      sheet = "PLNT18", skip = 1)
eGRID2020 <- read_excel("Data/Raw/eGRID2020_Data.xlsx",
                      sheet = "PLNT20", skip = 1)
Gov_Data <- read.csv("Data/Raw/US_Governors_1775_2020.csv")

States_shapefile_shp <- st_read("Data/Raw/us-state-boundaries.shp")


```
# Rationale and Research Questions

To conduct a preliminary exploration of the ways in which the political systems and the electricity grid coexist, our team has formulated the following research questions:

  1) What is the relationship between the percent penetration of renewable technology (i.e.,     solar, wind) and state governor (using the political affiliation of the state governor as     a proxy for a state’s political leanings) in a given year / over a period of time?
  
  2) Has renewable penetration grown in any states that have had a single-party governor over     a longer period of time?

\newpage

# Dataset Information

# Wrangling eGRID

```{r Wrangle eGRID}
# selecting data for desired columns from imported eGRID datasets

#Note column "PLPFGNCT" for 2010 -- note that "PLFUELCT" was dropped and replaced
eGRID2010_sub <- eGRID2010 %>%
  mutate(YEAR="2010") %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLPFGNCT,NAMEPCAP,LAT,LON)

eGRID2010_sub$PLFUELCT <- eGRID2010_sub$PLPFGNCT
eGRID2010_sub$PLPFGNCT <- NULL

#Note column "PLPFGNCT" for 2012 -- note that "PLFUELCT" was dropped and replaced
eGRID2012_sub <- eGRID2012 %>%
  mutate(YEAR="2012") %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLPFGNCT,NAMEPCAP,LAT,LON)

eGRID2012_sub$PLFUELCT <- eGRID2012_sub$PLPFGNCT
eGRID2012_sub$PLPFGNCT <- NULL

eGRID2014_sub <- eGRID2014 %>%
  mutate(YEAR="2014") %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLFUELCT,NAMEPCAP,LAT,LON)

eGRID2016_sub <- eGRID2016 %>%
  mutate(YEAR="2016") %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLFUELCT,NAMEPCAP,LAT,LON)

eGRID2018_sub <- eGRID2018 %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLFUELCT,NAMEPCAP,LAT,LON)

eGRID2020_sub <- eGRID2020 %>%
  select(YEAR,PSTATABB,ORISPL,PNAME,PLFUELCT,NAMEPCAP,LAT,LON)

# merging eGRID data, filtering for renewable fuels
eGRID_2010_2020 <- rbind(eGRID2010_sub,eGRID2012_sub,eGRID2014_sub,eGRID2016_sub,
      eGRID2018_sub,eGRID2020_sub)

# checking fuel types
unique(eGRID_2010_2020$PLFUELCT)

# filtering for desired fuel types
eGRID_2010_2020_RENEW <- filter(eGRID_2010_2020,PLFUELCT=="SOLAR"|PLFUELCT=="WIND"|PLFUELCT=="GEOTHERMAL"|
          PLFUELCT=="HYDRO"|PLFUELCT=="BIOMASS"|PLFUELCT=="NUCLEAR")

#Change year column to character
eGRID_2010_2020_RENEW$YEAR <- as.character(eGRID_2010_2020_RENEW$YEAR)

# checking fuel types were filtered correctly
unique(eGRID_2010_2020_RENEW$PLFUELCT)

# checking to ensure all years of interest are present
unique(eGRID_2010_2020_RENEW$YEAR)

# saving process data as CSV
write_excel_csv(eGRID_2010_2020_RENEW, 
                path = "Data/Processed/eGRID_2010_2020_RENEW.csv")
```
\newpage

#Wrangling Gov Data

```{r Wrangling Gov Data, include=FALSE}

#Filtering governor data for 2010 to 2020
Gov_Data_10_20 <- filter(Gov_Data,year >= 2010)

#Creating State abbreviation column for gov data
Gov_Data_State <- Gov_Data_10_20 %>%
  mutate(StateAbbreviation = state.abb[match(state, state.name)])

#Changing year to character
Gov_Data_State <- Gov_Data_State %>%
  mutate(year = as.character(year))

#Separating time in office column
#Key assumption is that end year is the year to be used for analysis of political
#Rationale -- incumbent governor is in office for the majority of this year
Gov_Data_State <- Gov_Data_State %>%
  separate(time_in_office, into = c("Start_Year", "End_Year"), sep = " - ")

# filtering for columns of interest
States_shapefile_shp_filtered <- States_shapefile_shp %>%
  select(basename, stusab, geometry)

#Merging governor data with state shape file
GOV_States <- left_join(Gov_Data_State, States_shapefile_shp_filtered, by = c("StateAbbreviation" = "stusab"))

#Joining gov data with eGRID processed data
GOV_eGRID_10_20 <- left_join(eGRID_2010_2020_RENEW, Gov_Data_State, by = c("PSTATABB" = "StateAbbreviation", "YEAR" = "year"))

# dropping N/A's
GOV_eGRID_10_20 <- GOV_eGRID_10_20 %>%
  drop_na(LAT)

#Checking years of final dataset
unique(GOV_eGRID_10_20$YEAR)

#Checking fuel types of final dataset
unique(GOV_eGRID_10_20$PLFUELCT)

#Dataset information -- REVISIT THIS
dim(GOV_eGRID_10_20)
colnames(GOV_eGRID_10_20)

#Adding Final Processed file to processed folder
write_csv(GOV_eGRID_10_20, 
                path = "Data/Processed/GOV_eGRID_10_20.csv")

```

\newpage

# Exploratory Analysis 
```{r Exploratory Time Series Analysis, include=FALSE, fig.retina=2, dev='pdf'}

#Exploratory distribution analysis using box, point, and violin plot

ggplot(GOV_eGRID_10_20, aes(x = YEAR, y = NAMEPCAP, color= PLFUELCT)) +
  geom_point() +
  ylim(0, 5000)

ggplot(GOV_eGRID_10_20, aes(x = YEAR, y = NAMEPCAP)) +
  geom_boxplot() +
  ylim(0, 50)

ggplot(GOV_eGRID_10_20, aes(x = YEAR, y = NAMEPCAP)) +
  geom_violin() +
  ylim(0, 50)

Republican_Renewables <- GOV_eGRID_10_20 %>%
  filter(party == "Republican") %>%
  ggplot(aes(x = YEAR, y = NAMEPCAP, color = PLFUELCT)) +
  geom_point() +
  ylim(0, 5000)
plot(Republican_Renewables)

Democrat_Renewables <- GOV_eGRID_10_20 %>%
  filter(party == "Democrat") %>%
  ggplot(aes(x = YEAR, y = NAMEPCAP, color = PLFUELCT)) +
  geom_point() +
  ylim(0, 5000)
plot(Democrat_Renewables)

```

\newpage

# Analysis

```{r Regression Analysis, include=FALSE}

#WRANGLING
# creating subset of governor data to sum no. of democrats in office by state
Years_in_Office <- Gov_Data_State %>%
  mutate(dem_count = ifelse(party=="Democrat",1,0),
         rep_count=ifelse(party=="Republican",1,0))%>%
  group_by(StateAbbreviation)%>%
  summarise(total_dem_count = sum(dem_count),total_rep_count=sum(rep_count))

# aggregating nameplate capacity by state
GOV_eGRID_10_20_TOTAL_CAPACITY <- GOV_eGRID_10_20 %>%
  group_by(PSTATABB) %>%
  summarize(TOTAL_NAMEPCAP = sum(NAMEPCAP))

  
# joining datasets to create data set for regression analysis
Capacity_by_Party <-left_join(Years_in_Office,GOV_eGRID_10_20_TOTAL_CAPACITY,by = c("StateAbbreviation"="PSTATABB"))
                               
#SINGLE REGRESSION
# running simple linear regression of Democratic party on nameplate capacity
Regression_DemParty_by_Capacity <- lm(Capacity_by_Party$TOTAL_NAMEPCAP~Capacity_by_Party$total_dem_count)

summary(Regression_DemParty_by_Capacity)

#PEARSON CORRELATION
# running a correlation test
cor.test(Capacity_by_Party$TOTAL_NAMEPCAP,Capacity_by_Party$total_dem_count)

#SPEARMAN CORRELATION
#Test for correlative coefficient
cor(Capacity_by_Party$TOTAL_NAMEPCAP,Capacity_by_Party$total_dem_count,method="spearman")

#MULTIPLE REGRESSION
# trying multiple linear regression
MP_Reg_Capacity_by_Party <- lm(data=Capacity_by_Party,TOTAL_NAMEPCAP~total_dem_count+total_rep_count)

summary(MP_Reg_Capacity_by_Party)

# plotting regression - capacity by Dem
ggplot(Capacity_by_Party, aes(x=total_dem_count, y=TOTAL_NAMEPCAP)) +
  geom_point()


# plotting regression - capacity by Rep
ggplot(Capacity_by_Party, aes(x=total_rep_count, y=TOTAL_NAMEPCAP)) +
  geom_point()

#ANOVA
#ANOVA Test for party correlative effect on nameplate capacity
ANOVA_Capacity <- aov(data = Capacity_by_Party, TOTAL_NAMEPCAP ~ total_dem_count)
summary(ANOVA_Capacity)

```
\newpage

# Geospatial Analysis

```{r Geospatial Analysis, include=FALSE}

#EXPLORATORY
#Viewing State Boundaries
mapview(States_shapefile_shp)

#Viewing power plants
Gov_eGRID.sf <- GOV_eGRID_10_20 %>% 
  st_as_sf(coords = c('LON','LAT'), crs=4326) %>%
  filter(PSTATABB == "NC")

mapview(Gov_eGRID.sf, zcol="PLFUELCT")

#COMBINE SPATIAL
#Layering Boundaries and plants
GOV_States.sf <- st_as_sf(GOV_States) %>%
  filter(StateAbbreviation == "NC")

mapview(GOV_States.sf, zcol = "party") +
  mapview(Gov_eGRID.sf, zcol="PLFUELCT")

party_colors <- c("Democratic" = "blue",  # Assigning blue to Democratic
                  "Republican" = "red"  # Assigning red to Republican
                  )

party_map <- mapview(GOV_States.sf, zcol = "party", col.regions = party_colors) +
  mapview(Gov_eGRID.sf, zcol="PLFUELCT")
print(party_map)

```

```{r}
#Playing around with data cuts

GOV_States.sf.NC <- GOV_States.sf %>%
  filter(StateAbbreviation == "NC")


#mapview(GOV_States.sf.NC, zcol = "party", col.regions = party_colors_NC)

#mapview(GOV_States.sf.NC, subset = GOV_States.sf.NC$year == "2013", zcol = "party", col.regions = party_colors_NC)

#mapview(GOV_States.sf.NC, subset = GOV_States.sf.NC$year == "2010", zcol = "party", col.regions = party_colors_NC)

#Note that "GOV_States.sf.NC$End_Year" is a character
#mapview(GOV_States.sf.NC, subset = GOV_States.sf.NC$End_Year == "2013", zcol = "party", col.regions = party_colors_NC) +
 #mapview(Gov_eGRID.sf, subset = Gov_eGRID.sf$PSTATABB == "NC", subset = #Gov_eGRID.sf$PLFUELCT == "SOLAR")




#party_colors_NC <- c("Democrat" = "blue",
 #                    "Republican" = "red")

#GOV_States.sf.NC$party <- as.factor(GOV_States.sf.NC$party)
#GOV_States.sf.NC$End_Year <- as.numeric(GOV_States.sf.NC$End_Year)


#party_colors_NC <- mapviewColors(x=GOV_States.sf.NC, 
                                 #zcol="party",
                                 #colors = c("red", "blue"),
                                #at = unique(GOV_States.sf.NC$party))


#mapview(filter(GOV_States.sf.NC,End_Year == "2013"), zcol = "party", col.regions = party_colors_NC)

#mapview(GOV_States.sf.NC, subset = GOV_States.sf.NC$End_Year == "2016", zcol = "party", col.regions = party_colors_NC)

```

## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 



\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
